---
title: "COVID reference-mapping task"
author: "António Sousa (e-mail: <aggode@utu.fi>) - Elo lab (<https://elolab.utu.fi>)"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output: 
   html_document:
      toc: true 
      toc_float: true
      theme: united
      code_folding: hide
---

<br>

<br>

---

<br>

<br>

# Reference-mapping task

<br>

This single-cell RNA-seq reference-mapping task describes an example of reference-mapping a human COVID-19 peripheral blood mononuclear cells (PBMC) against a previously annotate human PBMC data set. The COVID data set was downloaded as `Seurat` object from [cziscience](https://cellxgene.cziscience.com) (see the R script `01_create_datasets.R`) and the reference previously annotated PBMC data set retrieved from `SeuratData` (v.`r packageVersion("SeuratData")`): 

   + reference (`ref`): 5,000 genes x 36,433 cells
   
   + query (`query`): 17,374 genes x 14,783 cells
      
<br>

The analyses performed in this notebook rely in the `Seurat` (v.`r packageVersion("Seurat")`) and `Azimuth` (v.`r packageVersion("Azimuth")`). 

```{r setup, include=FALSE}
## Set up configuration for R chunks
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
if ( ! "klippy" %in% row.names(installed.packages()) ) remotes::install_github("rlesur/klippy")
```

```{r klippy, echo=FALSE}
## Configure the layout of icon to copy to clipboard the R chunk codes
klippy::klippy(position = c('top', 'right'))
```

Import the main packages used in this notebook: `Seurat` (v.`r packageVersion("Seurat")`), `SeuratWrappers` (v.`r packageVersion("SeuratWrappers")` - integration wrappers for Seurat), `dplyr` (v.`r packageVersion("dplyr")` - wrangling data), `patchwork` (v.`r packageVersion("patchwork")` - visualization), `ComplexHeatmap` (v.`r packageVersion("ComplexHeatmap")` - heatmap), `Azimuth` (v.`r packageVersion("Azimuth")` - reference-mapping).

```{r packages, message=FALSE, warning=FALSE}
## Import packages
library("dplyr") # data wrangling
library("Seurat") # scRNA-seq analysis
library("Azimuth") # reference-mapping
library("patchwork") # viz
library("ComplexHeatmap") # heatmap
```

Create output directories to save intermediate results, figures, tables and R objects. 

```{r dirs}
## Output directories
res.dir <- file.path("../results", "covid_refmap_task", c("plots", "tables", "objects"))
for (folder in res.dir) if (!dir.exists(folder)) dir.create(path = folder, recursive = TRUE)
```

<br>

<br>

---

<br>

<br>

## (1) Import datasets 

_(5 min)_

>AIM: Import and explore the Seurat object data. 

<br>

Import the human PBMCs from COVID-19 patients (`query`) as well as the healthy human PBMCs (`ref`) as `Seurat` objects located at `data/covid.rds` and `data/pbmcref`.  

```{r import data}
## Import query and reference Seurat objects
data.dir <- "../data"

# Reference
ref <- readRDS(file = file.path(data.dir, "pbmcref.rds"))

# Query
query <- readRDS(file = file.path(data.dir, "covid.rds"))
```

Explore quickly the `Seurat` `query` and `ref` objects.

```{r explore Seurat objects}
## Explore Seurat objects
# Print Seurat object
query
ref

# Structure
#str(query)
#str(ref)

# Check meta.data
head(query@meta.data)
head(ref@meta.data)

# Check how different cell types are in the reference 
table(ref$celltype.l1)
table(ref$celltype.l2)

# Check no. of genes 
nrow(query)
nrow(ref)

# Check no. of cells 
ncol(query)
ncol(ref)
```

<br>

<br>

---

<br>

<br>

## (2) DimRed viz 

_(10 min)_

>AIM: Visualize data in the low dimensional space highlighting the different categorical variables of interest. 

<br>

Check the metadata of query and reference objects and choose the most interesting categorical variables to highlight into UMAP. 

```{r dimred}
## Dimensional reduction - visualization

## Reference 
colnames(ref@meta.data)
ref.vars <- c("celltype.l1", "celltype.l2")
ref.umap.plts <- lapply(X = ref.vars, function(x) {
  DimPlot(object = ref, reduction = "refUMAP", group.by = x, pt.size = 0.1, label = TRUE)
})

## Query
colnames(query@meta.data)[4] <- "disease_stage"
colnames(query@meta.data)
query.vars <- c("disease_stage", "donor_id", "timepoint", "Cell.group", "Cell.class_reannotated")
query.umap.plts <- lapply(X = query.vars, function(x) {
  DimPlot(object = query, reduction = "umap", group.by = x, pt.size = 0.1, label = TRUE)
})
```

Plot the categorical variables `celltype.l1` and `celltype.l2` for the reference below. 

```{r dimred - plot ref, fig.width = 9, fig.height = 4.5}
## Plot dimensional reductions for reference
(ref.umap.plts[[1]] + NoLegend()) + (ref.umap.plts[[2]] + NoLegend()) 
```

<br>

Plot the categorical variables `disease_stage`, `donor_id`, `timepoint`, `Cell.group`, `Cell.class_reannotated` for the query below. 

```{r dimred - plot query, fig.width = 15, fig.height = 4.5}
## Plot dimensional reductions for reference
(query.umap.plts[[1]]) + (query.umap.plts[[2]]) + (query.umap.plts[[3]]) 
```

```{r dimred - plot query 2, fig.width = 10, fig.height = 5}
## Plot dimensional reductions for reference
(query.umap.plts[[4]] + NoLegend()) + (query.umap.plts[[5]] + NoLegend()) 
```

<br>

<br>

---

<br>

<br>

## (3) Reference-mapping

_(10 min)_

>AIM: Annotate and project the query against the reference data set. 

<br>

Perform reference-mapping below by identifying anchors between the COVID versus healthy PBMCs data sets and transferring the labels of interest (`celltype.l1`, `celltype.l2`) from the previously annotated healthy data set into the query COVID data set with the functions `FindTransferAnchors()` and `MapQuery()`. Alternatively you can run the same procedure using the `Azimuth` function `RunAzimuth`. 

```{r refmap}
## Reference-mapping

# Set to TRUE in case you wanna run with 'Azimuth'
run.azimuth <- FALSE

if (run.azimuth) {
  query <- Azimuth::RunAzimuth(query, reference = "pbmcref")
} else {
  # Find anchors
  anchors <- FindTransferAnchors(
    reference = ref,
    query = query, 
    features = rownames(Loadings(ref[["refDR"]])),
    reference.reduction = "refDR",
    normalization.method = "SCT",
    dims = 1:50
  )
  # Transfer labels
  query <- MapQuery(
    anchorset = anchors,
    query = query,
    reference = ref,
    refdata = list(
      celltype.l1 = "celltype.l1",
      celltype.l2 = "celltype.l2"
    ),
    reduction.model = "refUMAP"
  )
  ## EXCEPTION (to project query onto reference UMAP): due to the Azimuth reference 'pbmcref'
  query <- Azimuth:::NNTransform(object = query, meta.data = ref[[]]) # from 'RunAzimuth()'
  query[["ref.umap"]] <- RunUMAP(object = query[["query_ref.nn"]], reduction.model = ref[["refUMAP"]], 
                                  reduction.key = "UMAP_", verbose = TRUE)
}

# Save query R object
saveRDS(object = query, file = file.path(res.dir[3], "query_refmap.rds"))
```

<br>

<br>

---

<br>

<br>

## (4) Assess reference-mapping

_(15 min)_

>AIM: Assess the accuracy of the reference-mapping task. 

<br>

Run the R chunk code below to compare the predicted cell type labels for the COVID data set against the healthy human PBMCs with the ground-truth cell labels. 

```{r assess refmap accuracy}
## Assessment of reference-mapping accuracy

# Plot UMAPs
refmap.plts <- list()
refmap.plts[[1]] <- DimPlot(ref, reduction = "refUMAP", group.by =  "celltype.l1", 
                            label = TRUE, pt.size = 0.1, alpha = 0.1) + NoLegend()
refmap.plts[[2]] <- DimPlot(ref, reduction = "refUMAP", group.by =  "celltype.l2", 
                            label = TRUE, pt.size = 0.1, alpha = 0.1) + NoLegend()
refmap.plts[[3]] <- DimPlot(query, reduction = "ref.umap", group.by =  "Cell.group", 
                            label = TRUE, pt.size = 0.1, alpha = 0.1) + NoLegend()
refmap.plts[[4]] <- DimPlot(query, reduction = "ref.umap", group.by =  "Cell.class_reannotated", 
                            label = TRUE, pt.size = 0.1, alpha = 0.1) +
  NoLegend()
refmap.plts[[5]] <- DimPlot(query, reduction = "ref.umap", group.by = "predicted.celltype.l1", 
                            label = TRUE, pt.size = 0.1, alpha = 0.1) +
  NoLegend()
refmap.plts[[6]] <- DimPlot(query, reduction = "ref.umap", group.by = "predicted.celltype.l2", 
                            label = TRUE, pt.size = 0.1, alpha = 0.1) +
  NoLegend()
refmap.plts[[7]] <- DimPlot(query, reduction = "ref.umap", group.by =  query.vars[1], 
                            label = TRUE, pt.size = 0.1, alpha = 0.1)
refmap.plts[[8]] <- DimPlot(query, reduction = "ref.umap", group.by = query.vars[2], 
                            label = TRUE, pt.size = 0.1, alpha = 0.1)
refmap.plts[[9]] <- DimPlot(query, reduction = "ref.umap", group.by = query.vars[3], 
                            label = TRUE, pt.size = 0.1, alpha = 0.1)

# Confusion matrices
celltype1xgroup <- table(query$predicted.celltype.l1, query$Cell.group)
celltype1xgroup <- celltype1xgroup %>% as.matrix.data.frame(.) %>% 
  `colnames<-`(colnames(celltype1xgroup)) %>% `row.names<-`(row.names(celltype1xgroup))
celltype1xclass <- table(query$predicted.celltype.l1, query$Cell.class_reannotated)
celltype1xclass <- celltype1xclass %>% as.matrix.data.frame(.) %>% 
  `colnames<-`(colnames(celltype1xclass)) %>% `row.names<-`(row.names(celltype1xclass))
celltype2xgroup <- table(query$predicted.celltype.l2, query$Cell.group)
celltype2xgroup <- celltype2xgroup %>% as.matrix.data.frame(.) %>% 
  `colnames<-`(colnames(celltype2xgroup)) %>% `row.names<-`(row.names(celltype2xgroup))
celltype2xclass <- table(query$predicted.celltype.l2, query$Cell.class_reannotated)
celltype2xclass <- celltype2xclass %>% as.matrix.data.frame(.) %>% 
  `colnames<-`(colnames(celltype2xclass)) %>% `row.names<-`(row.names(celltype2xclass))
```

Plot below the heatmaps of the confusion matrices between predicted cell type labels (`celltype.l1`, `celltype.l2`) versus ground-truth cell type labels (`Cell.group`, `class_reannotated`).

```{r assess refmap accuracy - confusion matrices, fig.width=14, fig.height=7}
tbls <- list("celltype1xCell.group" = celltype1xgroup, "celltype1xclass_reannotated" = celltype1xclass, 
             "celltype2xCell.group" = celltype2xgroup, "celltype2xclass_reannotated" = celltype2xclass)
heat.list <- lapply(X = setNames(names(tbls), names(tbls)), FUN = function(comp) {
  Heatmap(matrix = t(apply(tbls[[comp]], 1, function(x) x/sum(x)*100)), name = "% of cells", 
          cluster_rows = F, cluster_columns = F, row_names_side = "left",
          show_column_names = T, show_row_names = TRUE,
          col = circlize::colorRamp2(c(0, 50, 100), c("white", "red1", "red4")), 
          column_names_side = "top", column_names_rot = 45, 
          layer_fun = function(j, i, x, y, width, height, fill, slice_r, slice_c) {
            v = pindex(tbls[[comp]], i, j)
            grid.text(sprintf("%.0f", v), x, y, gp = gpar(fontsize = 10))
            if(slice_r != slice_c) {
              grid.rect(gp = gpar(lwd = 2, fill = "transparent"))
            }
          }, 
          column_title = gsub("x", " vs ", comp), 
          rect_gp = gpar(col = "white", lwd = 2)
  )
})

# Print below
heat.list$celltype1xCell.group + heat.list$celltype1xclass_reannotated
heat.list$celltype2xCell.group + heat.list$celltype2xclass_reannotated
```

<br>

Project the data below onto the reference UMAP highlighting the predicted (`celltype.l1`, `celltype.l2`) and ground-truth cell type labels (`Cell.group`, `class_reannotated`).

```{r assess refmap accuracy - umaps ref, fig.width=9, fig.height=4.5}
refmap.plts[[1]] + refmap.plts[[2]]
```

```{r assess refmap accuracy - umaps query ground-truth, fig.width=9, fig.height=4.5}
refmap.plts[[3]] + refmap.plts[[4]]
```

```{r assess refmap accuracy - umaps query predicted, fig.width=9, fig.height=4.5}
refmap.plts[[5]] + refmap.plts[[6]]
```

Plot the categorical variables `disease_stage`, `donor_id`, `timepoint` for the integrated query below. 

```{r assess refmap accuracy - umaps query predicted vars, fig.width = 15, fig.height = 4.5}
## Plot dimensional reductions for reference
refmap.plts[[7]] + refmap.plts[[8]] + refmap.plts[[9]] 
```

<br>

<br>

---

<br>

<br>

#### R packages used and respective versions

<br>

```{r References, message=FALSE, warning=FALSE, paged.print=FALSE}
## R packages and versions used in these analyses
sessionInfo()
```

<br>

<br>

---

<br>

<br>
